Constraints on Cosmic-Ray Particle Populations, Magnetic Fields, and Dark Matter from VHE Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS

### 2011-09-19 ms.tex
New first paragraph of the abstract

\begin{abstract}
Theoretical models of galaxy clusters suggest they may be sources of very high-energy (VHE)
gamma-ray emission. Observations of radio halos and relics in clusters indicate populations of
relativistic electrons, which are either injected into the intra-cluster medium and accelerated by
turbulent shocks or secondaries from interactions of cosmic-ray hadrons (protons and ions) with
the intra-cluster gas. Those interactions could also produce gamma rays through the decay of
neutral pions. Reported here are observations of the Coma cluster of galaxies with the VERITAS
array of imaging Cherenkov telescopes. The observations were made between March and May
2008. Radio observations at 1.41 GHz taken with the Green Bank Telescope and X-ray observations
with the ROSAT All-Sky survey was used to a priori select regions of interest in the field of view.
No significant VHE gamma-ray emission from the Coma cluster was detected by VERITAS. Several
emission regions are considered, including the cluster core as both unresolved (point-like) and
spatially extended, and a select few cluster members. The flux upper limits at the 99\% confidence
level measured are on the order of $2-5\times 10^{-8}$ ph. m$^{-2}$ s$^{-1}$, which is two orders
of magnitude higher than model-predicted minimum gamma-ray fluxes.

The gamma-ray flux upper limits have been used to constrain the populations of cosmic-ray particles
and magnetic fields and in the Coma cluster of galaxies. The relative cosmic-ray pressure,
$X_{\mathrm{CR}}=\expval{P_{\CR}}/\expval{P_{\rmn{th}}}$, is on the order of 10-20\% when assuming
a gamma-ray photon index $\alpha=2.3$. The observations imply a lower limit on the intra-cluster
magnetic field on the order of a $\mu$G, depending on the magnetic decline and gamma-ray spectral
index. Finally, a constraint on the thermally averaged product of the total self-annihilation
cross section and the relative velocity of the dark-matter particles, $\expval{\sigma v}$,
is derived and found to be $\lesssim 10^{-21}$ cm$^{3}$ s$^{-1}$ for $m_{\chi}\gtrsim 1000$
GeV/c$^{2}$.
\end{abstract}

@@

\begin{deluxetable}{ccccccc}
\tablecolumns{7}
\tablewidth{0pc}
\tablecaption{Minimum gamma-ray fluxes in the hadronic model of radio halos, where the
radio-emitting electrons are secondaries from CR interactions and corresponding minimum relative
CR pressures for Coma} \rotate
\tabletypesize{\small}
\tablehead{
\colhead{RoI\tablenotemark{a} [deg]} 
& \multicolumn{3}{c}{$F_{\gamma,\,\rmn{min}}(>E)$\tablenotemark{b}} 
& \multicolumn{3}{c}{$X_{\CR,\,\rmn{min}}~[\%]$\tablenotemark{c} } \\
& \colhead{$\alpha=2.1$} & \colhead{$\alpha=2.3$} & \colhead{$\alpha=2.5$}
& \colhead{$\alpha=2.1$} & \colhead{$\alpha=2.3$} & \colhead{$\alpha=2.5$}
}
\startdata
\multicolumn{7}{c}{VERITAS energy range, $E>220$~GeV:}\\
\hline
0.13 & 1.6 & 0.7 & 0.3 & 0.067 & 0.061 & 0.11 \\
0.2  & 3.1 & 1.4 & 0.6 & 0.078 & 0.072 & 0.13 \\
0.4  & 6.3 & 2.8 & 1.3 & 0.098 & 0.090 & 0.16 \\
\cutinhead{{\em Fermi} energy range, 1--3~GeV:}
0.13 & ~~3.5 & ~~4.8 & ~~6.4 & 0.067 & 0.061 & 0.11 \\
0.2  & ~~6.8 & ~~9.3 &  12.5 & 0.078 & 0.072 & 0.13 \\
0.4  &  13.5 & 18.6  &  25.0 & 0.098 & 0.090 & 0.16 \\
\enddata
\tablenotetext{a}{Intrinsic source radius; we approximate the aperture of the point source-limits by the
analysis PSF of 0.13$^\circ$.}
\tablenotetext{b}{Minimum gamma-ray flux of the hadronic model described in Sect~\ref{sec:Fmin}. 
Values are in units of $10^{-10}$ ph. m$^{-2}$ s$^{-1}$ (VERITAS energy range) and $10^{-8}$ ph. m$^{-2}$
s$^{-1}$ {\em (Fermi) energy range}.}
\tablenotetext{c}{Minimum relative CR pressure, $X_{\CR,\,\rmn{min}}$, in the
hadronic model described in Sect~\ref{sec:Fmin}.}
\label{table:constraints}
\end{deluxetable}

>>>>

\begin{abstract}
Theoretical models of galaxy clusters suggest they may be sources of very high-energy (VHE)
gamma-ray emission. Observations of radio halos and relics in clusters are indicative efficient
acceleration processes and the gamma-ray emission is the result of highly relativistic
cosmic-ray protons and ions interacting with intracluster gas and magnetic fields. Reported
here are VHE gamma-ray observations of the Coma cluster of galaxies with the VERITAS array
of imaging Cherenkov telescopes, with complementing Fermi Gamma-ray Space Telescope
observations at GeV energies. No significant gamma-ray emission from the Coma cluster
was detected. Flux upper limits at the 99\% confidence level were measured to be on the order of $2-5\times 10^{-8}$ ph. m$^{-2}$ s$^{-1}$ (VERITAS, $>$220 GeV) and
$\sim 3\times 10^{-6}$ ph. m$^{-2}$ s$^{-1}$ ({\em Fermi}, 1-3 GeV), respectively. The limits are
two orders of magnitude higher than model-predicted minimum gamma-ray fluxes, which
assumes the radio-emitting electrons to be all secondaries from cosmic-ray interactions.

The gamma-ray flux upper limits have been used to constrain the populations of cosmic-ray
particles and magnetic fields and in the Coma cluster of galaxies. Using a simplified analytical
approach, the relative cosmic-ray pressure, $X_{\mathrm{CR}}=\expval{P_{\CR}}/\expval{P_{\rmn{th}}}$,
is constrained to be smaller than 20\% from VERITAS data and smaller than 4\% from
{\em Fermi} data. 
The observations imply a lower limit on
the intra-cluster magnetic field on the order of a $\mu$G, depending on the magnetic decline
and gamma-ray spectral index. Finally, a constraint on the thermally averaged product of the total self-annihilation
cross section and the relative velocity of the dark-matter particles, $\expval{\sigma v}$,
is derived and found to be $\lesssim 10^{-21}$ cm$^{3}$ s$^{-1}$ for $m_{\chi}\gtrsim 1000$
GeV/c$^{2}$.
\end{abstract}

@@

\begin{deluxetable}{ccccccc}
\tablecolumns{7}
\tablewidth{0pc}
\tablecaption{Minimum gamma-ray fluxes in the hadronic model of radio halos, where the
radio-emitting electrons are secondaries from CR interactions and corresponding minimum relative
CR pressures for Coma}
\rotate
\tabletypesize{\small}
\tablehead{
\colhead{RoI\tablenotemark{a} [deg]} 
& \multicolumn{3}{c}{$F_{\gamma,\,\rmn{min}}(>E)$\tablenotemark{b}} 
& \multicolumn{3}{c}{$X_{\CR,\,\rmn{min}}~[\%]$\tablenotemark{c} } \\
& \colhead{$\alpha=2.1$} & \colhead{$\alpha=2.3$} & \colhead{$\alpha=2.5$}
& \colhead{$\alpha=2.1$} & \colhead{$\alpha=2.3$} & \colhead{$\alpha=2.5$}
}
\startdata
\multicolumn{7}{c}{VERITAS energy range, $E>220$~GeV:}\\
\hline
0.13 & 1.6 & 0.7 & 0.3 & 0.067 & 0.061 & 0.11 \\
0.2  & 3.1 & 1.4 & 0.6 & 0.078 & 0.072 & 0.13 \\
0.4  & 6.3 & 2.8 & 1.3 & 0.098 & 0.090 & 0.16 \\
\cutinhead{{\em Fermi} energy range, 1--3~GeV:}
0.13 & ~~3.5 & ~~4.8 & ~~6.4 & 0.067 & 0.061 & 0.11 \\
0.2  & ~~6.8 & ~~9.3 &  12.5 & 0.078 & 0.072 & 0.13 \\
0.4  &  13.5 & 18.6  &  25.0 & 0.098 & 0.090 & 0.16 \\
\enddata
\tablenotetext{a}{Intrinsic source radius; we approximate the aperture of the point source-limits by the
analysis PSF of 0.13$^\circ$.}
\tablenotetext{b}{Minimum gamma-ray flux of the hadronic model described in Sect~\ref{sec:Fmin}. 
Values are in units of $10^{-10}$ ph. m$^{-2}$ s$^{-1}$ (VERITAS energy range) and $10^{-8}$ ph. m$^{-2}$
s$^{-1}$ ({\em Fermi} energy range).}
\tablenotetext{c}{Minimum relative CR pressure, $X_{\CR,\,\rmn{min}}$, in the
hadronic model described in Sect~\ref{sec:Fmin}.}
\label{table:constraints}
\end{deluxetable}
